---
title: "NY SID: Exploration of discharges and select utilization flags"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
linkcolor: blue
header-inlcudes:
    - usepackage{amsmath}
    - usepackage{float}
---


```{r rmd_setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
knitr::opts_chunk$set(include=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
knitr::opts_knit$set(root.dir='..')
```
```{r libraries}
library(haven)
library(tidyverse)
library(knitr)
library(kableExtra)
library(RColorBrewer)
```

```{r plot settings}
theme_set(theme_bw())

qual5 <- brewer.pal(5, 'Set2')
```

```{r functions}
############################################
sumStatsOutcome <- function(outcome, .data=data) {
	# Calculate summary statistics for one outcome by year for all payer types

	.data %>%
		select(year, !!outcome) %>%
		group_by(year) %>%
			summarise_at(.vars=vars(-group_cols()),
				.funs=list(
					mean=~mean(., na.rm=TRUE),
					sd=~sd(., na.rm=TRUE),
					q1=~quantile(., 0.25, na.rm=TRUE),
					q2=~quantile(., 0.50, na.rm=TRUE),
					q3=~quantile(., 0.75, na.rm=TRUE),
					min=~min(., na.rm=TRUE),
					max=~max(., na.rm=TRUE),
					NAs=~sum(is.na(.)))) %>%
			ungroup()
}


############################################
kblSummaryStats <- function(outcome, payer, sum_stats, ...) {
	# Kable output of summary statistics for one payer type for outcome

	if (all(grepl('_pr$', outcome))) {
		outcome_stub <- paste0(unique(gsub('[[:digit:]]_pr', '', outcome)), payer, '_pr_')
	} else {
		outcome_stub <- paste0(unique(gsub('[[:digit:]]', '', outcome)), payer, '_')
	}

	# Transpose tibble
	temp <- sum_stats %>%
		select(year, starts_with(outcome_stub)) %>%
		t()
	colnames(temp) <- temp[1,]
	rownames(temp) <- gsub(outcome_stub, '', rownames(temp))
	tbl_temp <- as.data.frame(temp[-1,])

	kable(tbl_temp, format='latex', booktabs=TRUE, ...) %>%
		kable_styling(latex_options=c('HOLD_position', 'striped'))
}

############################################
binScatter <- function(outcome, .data=data, geom='line', ylab, title) {
	# Scatter outcome vs binned HHI

	temp <- .data %>%
		select(year, bdtot, !!outcome) %>%
		gather(payer, y, !!outcome)

	ggplot(temp, aes(x=bdtot, y=y, col=payer)) +
		stat_summary_bin(fun.y='mean', na.rm=TRUE, bins=20
			, geom=geom) +
		scale_color_manual('Payer type', values=qual5, labels=c('Medicare', 'Medicaid', 'Private Insurance', 'Self-Pay', 'No charge')) +
		facet_wrap(~year, ncol=3) +
		labs(x='Beds', y=ylab, title=title) +
		theme(axis.text.x=element_text(size=6), axis.text.y=element_text(size=6),
			strip.text.x=element_text(size=6, margin=margin(t=0.6, b=0.6)),
			legend.title=element_text(size=6), legend.text=element_text(size=6), legend.key.size=unit(.5, 'line'), legend.position=c(.5, .15))
}
```

```{r data prep}
data <- read_dta('dump/hhi_ny_sid_supp_hosp.dta')

############################################
# Payer type labels
pay_labels <- c('Medicare', 'Medicaid', 'Private Insurance', 'Self-Pay', 'No charge')

# Outcome variables
ds_cnts <- paste0('discharges', 1:5)
ds_prs <- paste0(ds_cnts, '_pr')
ed_cnts <- paste0('u_ed', 1:5)
ed_prs <- paste0(ed_cnts, '_pr')
mhsa_cnts <- paste0('u_mhsa', 1:5)
mhsa_prs <- paste0(mhsa_cnts, '_pr')
newbn_cnts <- paste0('u_newbn', 1:5)
newbn_prs <- paste0(newbn_cnts, '_pr')

outcomes <- list(ds_cnts, ds_prs, ed_cnts, ed_prs, mhsa_cnts, mhsa_prs, newbn_cnts, newbn_prs)
```


# Table of contents

1. [Exploration of NY SID outcomes](#exploration-of-ny-sid-outcomes)
	a. [Distributions of outcomes by year](#distributions-of-outcomes-by-year)
	b. [Summary statistics of outcomes](#summary-statistics-of-outcomes)
	c. [Scatterplots of outcomes vs beds](#scatterplots-of-outcomes-vs-beds)
2. [Distribution of DRG codes within service utilization flags](#distribution-of-drg-codes-within-service-utilization-flags)



# Exploration of NY SID outcomes


## Distributions of outcomes by year
*Note:* unit of analysis is at hospital level

```{r outcome distributions, include=TRUE, results='asis', message=FALSE}
outcome_titles <- c('Discharges (counts)', 'Discharges (prop.)', 'Emergency Dept. (counts)', 'Emergency Dept. (prop.)', 'Mental Health/Substance Abuse (counts)', 'Mental Health/Substance Abuse (prop.)', 'Newborn (counts)', 'Newborn (prop.)')

# Loop through outcomes
for (i in 1:length(outcomes)) {
	outcome <- outcomes[[i]]

	# Loop through payer types
	for (pay in 1:5) {
		y <- outcome[[pay]]
		title <- paste0('Histogram of ', outcome_titles[[i]], ' - ', pay_labels[[pay]])

		print(ggplot(data, aes(get(y))) +
			geom_histogram(bins=20, na.rm=TRUE, col='black', fill='grey80') +
			facet_wrap(~year, ncol=3) +
			labs(x='', y='', title=title) +
			theme(strip.text.x=element_text(size=6, margin=margin(t=0.6, b=0.6))))
	}
}
```


## Summary statistics of outcomes
*Note:* in all tables, unit of analysis is at hospital level

```{r summary statistics, include=TRUE, results='asis'}
outcome_digits <- rep(c(0,2), times=length(outcomes)/2)
outcome_captions <- c('Discharges (counts)', 'Discharges (prop.)', 'Emergency Dept. (counts)', 'Emergency Dept. (prop.)', 'Mental Health/Substance Abuse (counts)', 'Mental Health/Substance Abuse (prop.)', 'Newborn (counts)', 'Newborn (prop.)')

# Loop through outcomes
outcome_summaries <- lapply(outcomes, sumStatsOutcome)

for (i in 1:length(outcomes)) {
	outcome <- outcomes[[i]]
	sum_stats <- outcome_summaries[[i]]
	digits <- outcome_digits[[i]]
	
	# Loop through payer types
	for (pay in 1:5) {
		caption <- paste0(outcome_captions[[i]], ' - ', pay_labels[[pay]])
		print(kblSummaryStats(outcome, pay, sum_stats, digits=digits, caption=caption))
	}
}
```


## Scatterplots of outcomes vs beds
*Note:* in all plots, beds are binned (bins=20) on the x-axis

```{r scatter outcomes, include=TRUE, results='asis', message=FALSE, fig.fullwidth=TRUE}
outcome_ylabs <- rep(c('Counts', 'Prop.'), times=length(outcomes)/2)
outcome_titles <- c('Discharges (counts)', 'Discharges (prop.)', 'Emergency Dept. (counts)', 'Emergency Dept. (prop.)', 'Mental Health/Substance Abuse (counts)', 'Mental Health/Substance Abuse (prop.)', 'Newborn (counts)', 'Newborn (prop.)')

# Loop through outcomes
for (i in 1:length(outcomes)) {
	outcome <- outcomes[[i]]
	ylab <- outcome_ylabs[[i]]
	title <- outcome_titles[[i]]

	print(binScatter(outcome, ylab=ylab, title=title))
}
```


# Distribution of DRG codes within service utilization flags
